---
title: "Financial situation is greatest obstacle faced by Canadian students"
author: "Rebecca Yang"
date: '2020-09-25'
header-includes:
  - \usepackage{booktabs}
  - \usepackage{subfig}
slug: financial-situation-is-greatest-obstacle
categories: []
tags: []
output: blogdown::html_page


---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(kableExtra)
mycols <- c("seagreen2", "dodgerblue","orchid2", "slateblue2",  "darkturquoise", "salmon1") # my color palette
```


# Abstract 
Many Canadian students have aspirations for higher levels of education, but do not always believe that it is likely to achieve them. This research studied the obstacles they face in obtaining their desired level of education as reported in the 2016 General Social Survey conducted by Statistics Canada. It found that financial situation was the most common obstacle and the majority of obstacles were reported by women. These findings reiterate the existence of roadblocks in attaining education and the importance for government funding and financial aid to make education accessible for all students. 

# Introduction 

First paragraph: General. Motivational. Majors specific finding. 

"Stay in school, kids". This mantra has been preached to children for generations, instilling in them the idea that education is the key to social mobility, financial security, and a stable future. This holds even more true for post-secondary education; the 2016 Canadian census reported that men and women with a bachelor’s degree earned 47.2% and 58% more, respectively, than their counterparts with only high school diplomas (Statistics Canada). However, while society has made education very desirable, the question of whether or not it is attainable for all students remains. Perhaps unsurprisingly, the most commonly identified obstacle in getting education was the respondent's financial situation. 

Second paragraph: More about what you did and found.


Many students reported a difference in their desired level of education and the level of education they believed they were likely to complete. The disconnect between students’ aspirations and what they believed possible warranted further investigation into what exactly is standing between them and their desired education. After all, since education is such a good thing, shouldn't students be able to get all the education they want? When asked if they faced obstacles in obtaining education, almost a third of students responded "Yes". Of those students, 68% identified financial situation as the leading obstacle. It was also found that the majority of respondents who reported facing obstacles were women over men. The breakdown of the specific obstacles faced varied between sexes and age groups as well. 

Some weaknesses of these findings should be acknowledged. The data used was from a survey that was targeted towards all Canadians, so students are only a subset of the full sample. Consequently, when the data was broken down by the different obstacles and compared between different groups, the sample sizes became very small (n<30). As such, these findings should be taken with a grain of salt and merely serve as the impetus for further discussion and investigation. A survey targeted towards students about the obstacles they face and ways they can be best supported is the logical next step in bridging the gap between students’ educational hopes and realities. 



It seems counterintuitive to promote education as a society while still having such high barriers to postsecondary education. 



# Data

The data used is from the General Social Survey (GSS), a survey conducted annually by Statistics Canada in order to gain insight into Canadians’ social lives, habits, views, and well-being, and usually focuses on a particular aspect each year. This analysis makes use of the 2016 iteration of the GSS, which focused on “Canadians at Work and Home”. The target population was residents of Canada, except for the Yukon, Northwest Territories, and Nunavut, aged 15 and over that were not institutionalized.

Individuals who reported that their main activity was going to school were identified as students and asked specific questions about their educational plans and goals. Non-college certificates, private business programs, trade school programs, and the like were considered together as "Non-college whatever". Students over the age of 25 were also considered together as one age group. 

The variables of interest were the following:

* likely_future_edu: the level of education the respondent thought they were likely to attain
* desired_future_edu: the level of education the respondent desired to attain
* sex
* age_group
* obstacles: whether or not the respondent faced obstacles in attaining their desired education
* obstacles_future_edu: the specific obstacles identified by the respondents   

Summary statistics for these variables...
```{r, include= FALSE }
# tables 

student_data <- read_csv("student_data.csv")  %>% mutate(age_group = ifelse(age_group == "15 to 24 years", "15 to 24 years", "25 and over")) %>% mutate(obstacles_future_edu = ifelse(obstacles_future_edu == "Other - Specify", "Other", obstacles_future_edu)) # loading data

sex_table <- student_data %>% group_by(sex) %>% count() %>% mutate(prop=paste(round(n/nrow(student_data), 2)*100, "%")) 

age_table <- student_data %>% group_by(age_group) %>% count() %>% mutate(prop=paste(round(n/nrow(student_data), 2)*100, "%")) 

obstacles_data <- student_data %>% filter(obstacles != "NA") 
obstacles_table <- obstacles_data %>%  group_by(obstacles) %>% summarise(n=n()) %>% mutate(prop=paste(round(n/nrow(student_data), 2)*100, "%"))  

what_obstacles_data <- student_data %>% filter(obstacles_future_edu!= "NA") 
what_obstacles_table <- what_obstacles_data %>% group_by(obstacles_future_edu)%>% summarise(n=n()) %>% mutate(prop=paste(round(n/nrow(what_obstacles_data), 2)*100, "%")) 

big_table <- rbind(sex_table, (age_table %>% rename(sex = age_group)), (obstacles_table %>% rename(sex = obstacles)), (what_obstacles_table %>% rename(sex = obstacles_future_edu)))
 
c1 <- c("Sex", "Sex", "Age Group", "Age Group", rep("Obstacles", 9))
c2 <- c("Female", "Male", "15 to 24 years", "25 and over", "Don't know", "No", rep("Yes", 7))
c3 <- c("", "", "", "" , "", "", "", what_obstacles_table$obstacles_future_edu)

big <- cbind(c1, c2, c3, big_table$n, big_table$prop)
colnames(big) <- c("Variable", "", "", "Number of Respondents", "Proportion")
big_table <- kbl(big, caption = "Summary Statistics", align = c("lllrr")) %>% 
     collapse_rows(columns = 1:2, valign = "top") %>% kable_styling(bootstrap_options = c("condensed", "striped"))

```

```{r, echo= FALSE}
big_table
```



You must include graphs.
```{r, include=FALSE}
# graphs 

likely_students <- student_data %>% filter(likely_future_edu != "NA", desired_future_edu != "NA") %>% group_by(likely_future_edu) %>% summarise(n=n()) %>% mutate(future_edu = "likely") %>% rename(edu_level = likely_future_edu) %>% mutate(prop = n*100/nrow(student_data))

# desired education levels of students
desired_students <- student_data %>% filter(likely_future_edu != "NA", desired_future_edu != "NA") %>%  group_by(desired_future_edu) %>% summarise(n=n()) %>% mutate(future_edu = "desired") %>% rename(edu_level = desired_future_edu) %>% mutate(prop = 100*n/nrow(student_data))

future_education_data <- rbind(likely_students, desired_students)

# plot of proportions of students' desired and likely 

future_edu_props <-  future_education_data %>% ggplot(aes(y =reorder(edu_level, prop ), x=prop, fill = future_edu)) + geom_bar(stat = "identity", position = "dodge") + scale_fill_manual(values=mycols) + theme_minimal()  + labs(title = "Students' Desired vs. Likely Education Levels", x="Percentage of Responses (%)", y = "Education Level", fill = "")  + theme(plot.title = element_text(hjust = 0.5))  + geom_text(aes(label = paste(round(prop), "%")), size = 2.7, position = position_dodge(width = 1), hjust = -0.1) + scale_y_discrete( labels = c("Other", "High School Diploma/Equivalent", "Non-college Certificate", "Professional (i.e. MD)" , "Ph.D." , "College/CEGEP", "Master's", "Bachelor's" )) + scale_x_continuous(limits = c(0, 40))


```

```{r, echo = FALSE, fig.cap = "A comparison of students' desired and likely education levels"}
future_edu_props
```

The distinction between likely and desired education by the survey was of particular interest. 
The following graph compares the percentage of students who would like to attain a certain level to the percentage of students who believe that they will actually accomplish that level. From the graph, it can be seen that Master’s and Ph.D. degrees are not perceived to be attainable, as not all students that desired them believed they were likely to acquire them. Instead, they seem to have resigned themselves to the fact that they will probably "settle" for a Bachelor’s, as 35% of students think they will likely get a Bachelor’s compared to 27% that are actually aiming for them. On the flip side, for lower levels such as college degrees, non-college certificates, and high school diplomas, smaller gaps exist. 


The disconnect between students’ aspirations and what they believed possible warranted further investigation into what stands between them and their desired education. 

When asked if they faced obstacles in obtaining education, 31% of students responded “Yes”. 
Financial situation was overwhelmingly identified as the main obstacle students faced, comprising 68% of responses. Other obstacles identified included the length of time the program requires (5%), the inability to get into a program (6%), personal health (5%), and having to care for children (4%). 
```{r, include = FALSE}
future_obstacles <- student_data %>% filter(obstacles != "NA") %>% group_by(obstacles) %>%summarise(n = n(), prop = n*100/nrow(student_data)) %>%  ggplot(aes(x = reorder(obstacles, -prop), y=prop)) + geom_bar(stat = "identity", position = "dodge", fill = mycols[1:3])  + theme_minimal()  + labs(x = "Response", y = "Percentage of Responses (%)") + ggtitle("Do You Face Obstacles in Obtaining Your Desired Education?") + theme(plot.title = element_text(hjust = 0.5))  + geom_text(aes(label = paste(round(prop), "%")), size = 4, vjust = -0.25) + scale_y_continuous(limits = c(0,100))

what_obstacles <- what_obstacles_data %>% mutate(obstacles_future_edu = ifelse(obstacles_future_edu == "Other - Specify", "Other", obstacles_future_edu)) %>% group_by(obstacles_future_edu) %>% summarise(prop = n()*100/nrow(what_obstacles_data)) 

# specific obstacles faced 

what_obstacles <- what_obstacles %>% ggplot(aes(y = obstacles_future_edu, x = prop)) + geom_bar(stat = "identity", position = "dodge", fill = mycols)+ theme_minimal() + labs(y = "Obstacles", x = "Percentage of Responses (%)") + ggtitle("Obstacles Faced in Obtaining Desired Education") + theme(plot.title = element_text(hjust = 0.5))  + geom_text(aes(label = paste(round(prop), "%")), size = 3.5, hjust = -0.2) + scale_x_continuous(limits = c(0, 80))

```

```{r, echo = FALSE, fig.width = 2.5, fig.cap = "Students were asked if they faced obstacles in obtaining education"}
future_obstacles

```

```{r, echo = FALSE, fig.cap = "Students were asked to identity the obstacles they faced "}
what_obstacles
```

```{r, include = FALSE}
obs_n <- nrow(obstacles_data)
obstacles_students <-  student_data %>% filter(obstacles== "Yes")

fem<- nrow(obstacles_students %>% filter(sex == "Female", obstacles_future_edu != "NA"))
male <- nrow(obstacles_students %>% filter(sex == "Male", obstacles_future_edu != "NA"))
sex <- obstacles_students %>% filter(obstacles_future_edu != "NA") %>% group_by(sex, obstacles_future_edu) %>% count() %>% summarise(prop = ifelse(sex=="Male", 100*n/male, 100*n/fem))

obstacles_sex <- sex %>% ggplot(aes(y=reorder(obstacles_future_edu, prop), x = prop, fill = sex)) + geom_bar(stat = "identity", position = "dodge") + theme_minimal() + labs(title = "Do Men and Women Face the Same Obstacles?", y = "", x = "Percentage Affected (%)", fill = "Sex") + scale_fill_manual(values = c("orchid2", "dodgerblue")) + geom_text(aes(label = paste0(round(prop),"%")), position = position_dodge(width = 1), hjust= -0.25, size = 3.5) + scale_x_continuous(limits = c(0, 80))

```


This is unsurprising given the rising costs of higher education across Canada. According to Statistics Canada, in 2015, 50% of students graduated with some form of student debt. 


The finding that the financial burden of post-secondary education is an obstacle for students may seem like common knowledge, but evidently it is still not obvious to the most important people. In 2019, the Ontario provincial government announced cuts of $670 million to OSAP, the Ontario Student Assistance Program, which provides financial aid through loans and grants to lower-income students. While they justified it through a 10% tuition reduction, the reduction in grants was much more significant, forcing students to shoulder more crippling student debt to finance their education. 


```{r, include = FALSE}
young<- nrow(obstacles_students %>% filter(age_group == "15 to 24 years", obstacles_future_edu != "NA"))
old <- nrow(obstacles_students %>% filter(age_group  == "25 and over", obstacles_future_edu != "NA"))
age <- obstacles_students %>% filter(obstacles_future_edu != "NA") %>% group_by(age_group, obstacles_future_edu) %>% count() %>% summarise(prop = ifelse(age_group == "15 to 24 years", 100*n/young, 100*n/old))

obstacles_age <- age %>% ggplot(aes(y=reorder(obstacles_future_edu, prop), x = prop, fill = age_group)) + geom_bar(stat = "identity", position = "dodge") + theme_minimal() + labs(title = "Do Different Age Groups Face the Same Obstacles?", y = "", x = "Percentage Affected (%)", fill = "Age Group") + scale_fill_manual(values = mycols[4:5]) + geom_text(aes(label = paste0(round(prop),"%")), position = position_dodge(width = 1), hjust= -0.25, size = 3.5) + scale_x_continuous(limits = c(0, 80))
obstacles_age

```

Breaking down the obstacles faced by gender found that the majority of obstacles were reported by women. In particular, child care had the greatest disproportionate effect. Of those who said caring for children was an obstacle, a 93% of them were women, while women only made up 54% of the sample. Similarly, caring for children was an issue that primarily affected students aged 25 and over. 
```{r, echo = FALSE, fig.cap = "Women are more likely to face child care as an obstacle"}
obstacles_sex  
```
```{r, echo = FALSE, fig.cap = "Child care is a greater issue for older students"}
obstacles_age
```

# Discussion

My discussion... main findings of your graph.


Features of your survey.... Non-response... etc. And how these could affect the results....

Since the target population was all Canadians, not specifically students, it seems likely that mature students (over age 25) were overrepresented in the sample. This may have influenced the results as people of different ages have different priorities and challenges in their respective lives. For example, financial situation was an obstacle for 68% of students aged 15-24, but only affected 50% of students aged 55-64. Child care was practically a nonexistent issue for students aged 15-45, but was an obstacle for 21% of 35-44 year olds. However, this can also be seen as a  benefit of this survey as it incorporates a portion of the student population that may otherwise be overlooked, instead of focusing on the  “typical” student. Of course, this depends on whether the aim is to help younger students. Education may be considered a better investment for younger students than students aged 55-65, for example. 



## Weaknesses and next steps
```{r, include = FALSE}
total_table <- what_obstacles_data %>%  group_by(obstacles_future_edu) %>% summarise(n = n())
female_table <- what_obstacles_data %>% filter(sex == "Female") %>%  group_by(obstacles_future_edu) %>% summarise(Female = n())
male_table <- what_obstacles_data %>% filter(sex == "Male") %>%  group_by(obstacles_future_edu) %>% summarise(Male = n())
sex_table_obstacles  <- cbind(female_table, Male = male_table$Male, Total = total_table$n)
colnames(sex_table_obstacles ) <- c("Obstacles", "Female", "Male", "Total")
sex_table_obstacles  <- sex_table_obstacles  %>% kbl() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"), position = "left") %>% column_spec(2, color = c("red", "black", rep("red", 4))) %>% column_spec(3, color = c("red", "black", rep("red", 4))) %>% column_spec(4, color = c("red", "black", "red", "black", "red", "red"))

```

```{r, echo = FALSE}
sex_table_obstacles 
```

Given that students were not the target population of the survey and represent a small subset of the sample, there are obvious weaknesses with this analysis. After breaking down the data to focus on students who faced obstacles and their specific obstacles, the sample sizes became very small (n < 30). For example, there were only 15 respondents who identified caring for children as an obstacle.  As such, these findings should be taken with a grain of salt and merely serve as the impetus for further investigation and discussion. A survey targeted towards students about the obstacles they face and ways they can be best supported, perhaps with particular focus on women and childcare, is the logical next step in bridging the gap between students’ educational hopes and realities. 


# References
- Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software,
  4(43), 1686, https://doi.org/10.21105/joss.01686
- Statistics Canada. 2020. "Half of recent postsecondary graduates had student debt prior to the pandemic." The Daily. July 28. Statistics Canada Catalogue no. 11-001-XIE.
http://www.statcan.ca/Daily/English/040728/d040728a.htm (accessed August 16, 2005).
- Jones, A. (2019, September 1). Province's cuts to OSAP mean more debt, stress as students head back to school | CBC News. https://www.cbc.ca/news/canada/toronto/osap-debt-stress-students-1.5267350. 
- R Core Team (2020). R: A language and environment for statistical computing. R, Foundation for Statistical Computing, Vienna, Austria. URL, https://www.R-project.org/.
- JJ Allaire and Yihui Xie and Jonathan McPherson and Javier Luraschi and Kevin Ushey and Aron Atkins and Hadley Wickham and Joe Cheng and Winston Chang and Richard Iannone (2020). rmarkdown: Dynamic Documents for R. R package version 2.3. URL https://rmarkdown.rstudio.com.
- Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2020). dplyr: A
  Grammar of Data Manipulation. R package version 1.0.2.
  https://CRAN.R-project.org/package=dplyr
- Statistics Canada. 2017. General social survey on Canadians at Work and Home (cycle 30), 2016 (public-use microdata file). Statistics Canada (producer). Using CHASS (distributor).
https://sda-artsci-utoronto-ca.myaccess.library.utoronto.ca/cgi-bin/sda/hsda?harcsda4+gss30 (accessed September 25, 2020). Accompanying documentation: user guide and questionnaire (PDF). 
- Sam Firke (2020). janitor: Simple Tools for Examining and Cleaning Dirty Data. R
  package version 2.0.1. https://CRAN.R-project.org/package=janitor
-  Hao Zhu (2020). kableExtra: Construct Complex Table with 'kable' and Pipe Syntax. R
  package version 1.2.1. https://CRAN.R-project.org/package=kableExtra
- Alexander, Rohan (2020) gss_cleaning.R [Source code]. 
